---
title: "Chapter 4: Intro to Map Making"
output: 
  html_document:
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 5
    toc_float: 
      collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css style settings, echo = FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 15px;
    border-left: 5px solid #90e0ef;
}

.purplebox {
  padding: 1em;
  background: #f0e8fa;
  color: #5c24a1;
  border: 2px dotted #926dc0;
  border-radius: 4px;
}

.purple_border {
  padding: 1em;
  background: white;
  color: #5c24a1;
  border: 2px dotted #926dc0;
  border-radius: 4px;
}

.center {
  text-align: center;
}
.left {
  text-align: left;
}
```


### 4.1 Lab Goals 

This chapter aims to introduce basic map making concepts and techniques. It applies the analysis workflow that we have learnt in [Chapter 3](https://fuzhen-yin.github.io/uccs_geoviz/lab3.html) to spatial visualization. After completing this tutorial, you should be able to:

* **Understand** the basic mapping concepts in R;
* **Interpret** and **Manipulate** tabular and spatial data collected from U.S. Census;
  + data [cleaning]{.ul} 
  + data [calculation]{.ul}
  + [filtering]{.ul} records
  + [joining]{.ul} data by columns 
* **Create** and **Export** near publishable static maps;
  + producing [choropleth]{.ul} map
  + adding [legend]{.ul}, [scale bar]{.ul} and [north arrow]{.ul}
  + [combining]{.ul} multiple maps together


### 4.2 Good Practice

#### 4.2.1 Organizing Folders & Sub-folders
Similar to what we have done in previous labs, under the course folder, please create a folder called "**lab4**". Next, in the **lab4** folder, please create two [sub-folders]{.ul} that one is called "**data**" and another one is "**plot**". 
We'll use the "data" folder to store data collected from the internet. The "plot" will be used to export plots generated from R. 


#### 4.2.2 Data
In this lab, we continue to explore **US Census Bureau** as a potential data source for administrative boundaries, and demographic, economic and population data. Specifically, we can collect:

> [Cartographic Boundaries](https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html) by geography (e.g., cities, counties, census tracts, block groups) or by scale. And 

> [Census Tables](https://data.census.gov/) summarizing the socio-economic and demographic characteristics of particular areas.  

This practical will use **El Paso** county (where UCCS is located) as a case study to explore its **poverty level** through map making. Please follow the steps below to [download]{.ul} data, [unzip]{.ul} it and [move]{.ul} the data to the required folder. 

* **Go** to TO_UPDATE
* **Download** the file _"lab4_data.zip"_
* **Unzip** folder _"lab4_data.zip"_
* **Move** all files from the "_lab4_data_" folder to the _"data"_ folder under "_lab4_" see [Step 4.2.1](https://fuzhen-yin.github.io/uccs_geoviz/lab4.html#421_Organizing_Folders__Sub-folders) 


If there you have any questions about the above-mentioned steps, please refer to [Chapter 3.2.3](https://fuzhen-yin.github.io/uccs_geoviz/lab3.html#322_Data) for detailed instructions. 





### 4.3 Viz: Colorado Counties Map 

This section starts with creating a simple map showing all counties in the state of Colorado. 

Before that, let's load some packages and 




```{r, eval=FALSE}
#

```

### 4.4 


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```



```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```


```{r, eval=FALSE}
#

```





