---
title: "Chapter 5: Map Making (2) Spatial Data Manipulation"
output: 
  html_document:
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 5
    toc_float: 
      collapsed: true
css: website.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 5.1 Lab Goals

This chapter continues to explore the techniques and analyses related to map-making. In [Chapter 4](https://fuzhen-yin.github.io/uccs_geoviz/lab4.html), we have played with census data (e.g., tables and cartographic boundaries) to make choropleth maps. Today, we will use [Airbnb data](https://insideairbnb.com/get-the-data/) as an example to learn more advanced spatial analysis skills including:

* **Converting** .csv file to spatial data 
* **Spatial joining** 
* Mapping and analyzing **point** data
* Using **functions** in R 


### 5.2 Good Practice

#### 5.2.1 Organizing Folders & Sub-folders
Under the course folder, please create a folder called "**lab5**". Next, in the **lab5** folder, please create two [sub-folders]{.ul} that one is called "**data**" and another one is "**plot**". 

#### 5.2.2 Data
In this lab, we will explore data sources other than [US Census Bureau](https://data.census.gov/). Particularly, we will play with [Airbnb data](https://insideairbnb.com/get-the-data/) for the city of Denver (CO). 

> [AirBnb](https://insideairbnb.com/get-the-data/) has published datasets related to [listings]{.ul}, [reviews]{.ul}, and [neighborhoods]{.ul} in major cities or tourist destinations all over the world. [Here](https://docs.google.com/spreadsheets/d/1iWCNJcSutYqpULSQHlNyGInUvHg2BoUGoNRIGa6Szc4/edit?gid=1322284596#gid=1322284596) you can find the data dictionary of [Airbnb data](https://insideairbnb.com/get-the-data/). 



Please follow the steps below to [download]{.ul} data, [unzip]{.ul} it and [move]{.ul} the data to the required folder. 

* **Go** to https://github.com/fuzhen-yin/uccs_geoviz/blob/main/data/lab5_data.zip
* **Download** the file _"lab5_data.zip"_
* **Unzip** folder _"lab5_data.zip"_
* **Move** all files from the "_lab5_data_" folder to the _"data"_ folder under "_lab5_" see [Step 5.2.1](https://fuzhen-yin.github.io/uccs_geoviz/lab5.html#521_Organizing_Folders__Sub-folders) 


If there you have any questions about the above-mentioned steps, please refer to [Chapter 3.2.3](https://fuzhen-yin.github.io/uccs_geoviz/lab3.html#322_Data) for detailed instructions. 


#### 5.2.3 Launching R Studio

Again, we would like to start a **new project** from scratch with a **clean R Script**. Please do the following steps. If you have any questions about these steps, please refer to the relevant chapters for help. 

* Step 1: Make sure all existing R projects are **properly closed**. 
    + If *not*, please close it by going to *File* --> *Close Project* --> *Save* changes (see [Chapter 2.5](https://fuzhen-yin.github.io/uccs_geoviz/lab2.html#25_Close__Exit)). 
* Step 2: Create a **New Project** using **Existing Directory**, navigate to **lab5**, click **open**, then **Create Project**. (see [Chapter 1.3](https://fuzhen-yin.github.io/uccs_geoviz/lab1.html#13_Launching_RStudio)). 
* Step 3: Create a **New Script** by go to **File** --> **New File** --> **R Script**. **Save** the script by giving it a proper name.


#### 5.2.4 Before Start 

:::: {.greenbox data-latex=""}
::: {.left data-latex=""}
[**Heads-Up! **]{.ul}
:::

All scripts is **non-copyable**. Here are some **coding standards** and **best practices** to follow: 

* Use **comments** to explain code (try to annotate the codes by yourself). 
* Ensure there is sufficient **whitespace between lines** of text to improve legibility. 
* **Consistent** variable naming
* When naming a new column, creating a new folder or file, please:
    + **Avoid special characters** such as ", * _ = & " .
    + **Avoid adding blank spaces**. For example, instead of using a name as "plot 1" (x), please use "plot_1" (✓). 

::::


### 5.3 Viz: Mapping Airbnb Listings in Denver 

We start with mapping Airbnb listings. Unfortunately, there is no [Airbnb data](https://insideairbnb.com/get-the-data/) for Colorado Springs. But it has data for [Denver]{https://insideairbnb.com/denver/}. 

Load libraries. We have seen most of these libraries except for ``` geojsonsf ``` which is often used to read spatial data in [.geojson]{.ul} format and convert it to simple features. 

#### 5.3.1 Libraries & Data
```{r, eval=FALSE, class.source = "nocopy"}
## install geojsonsf
install.packages("geojsonsf")

# library 
library(sf)         # simple features 
library(tmap)       # mapping
library(ggplot2)    # plot
library(dplyr)      # data manipulation
library(geojsonsf)  # geojson
```


Read data  
```{r, eval=FALSE, class.source = "nocopy"}
# US Cartographic Boundaries: Cities & towns in Colorado 
co_place <- st_read("data/cb_2023_08_CO_place_500k/cb_2023_08_place_500k.shp")


# Airbnb data 
ab_listing <- read.csv("data/airbnb_denver/listings.csv")
ab_nghbr <- geojson_sf("data/airbnb_denver/neighborhood.geojson")
```

Examine data. Please have a look at these tables, and report: 

**[Q1]**: How many records in "co_place"? <br>
**[Q2]**: Which columns could be used to extract "Denver" from "co_place"? <br>
**[Q3]**: In "ab_nghbr", how many records (listings)? <br>
**[Q4]**: In "ab_nghbr", which two columns contain spatial information and could be used for mapping?  <br>

```{r, eval=FALSE, class.source = "nocopy"}
View(co_place)
View(ab_nghbr)
View(ab_listing)
```

#### 5.3.2 Exploratory Analysis of Airbnb Listing 

Check columns information in "ab_nghbr"
```{r, eval=FALSE, class.source = "nocopy"}
# column info
str(ab_listing)
```

For all listings in Denver, count different [room types]{.ul}. The function ``` table() ``` builds a contingency table.   

**[Q6]**: Report the number of listings of different room types. <br>
**[Q7]**: Provide a screenshot of "plot1_airbnb_den_listing_room_type.pdf". 
```{r, eval=FALSE, class.source = "nocopy"}
# count room types
table(ab_listing$room_type)

# quick plot of room types 
pdf("plot/plot1_airbnb_den_listing_room_type.pdf")
table(ab_listing$room_type) %>% barplot()
dev.off()
```

Descriptive statistics listings' prices. 
```{r, eval=FALSE, class.source = "nocopy"}
# statistics
summary(ab_listing$price)
```

#### 5.3.3 Create Spatial Data from Table 

Check object class of "ab_listing"
```{r, eval=FALSE, class.source = "nocopy"}
# class
class(ab_listing)
```

"ab_listing" is a data frame. The two columns ``` longitude ``` and ``` latitude ``` contain spatial information of listings. Let's use these two columns to convert "ab_listing" to spatial data (simple features) and save it as a new variable ``` sf_ab_listing ```.


**[Q8]**: Compared to the data frame "ab_listing", which new column has been added to "sf_ab_listing"?
```{r, eval=FALSE, class.source = "nocopy"}
# table to simple features
sf_ab_listing <- st_as_sf(ab_listing, 
                          coords = c("longitude", "latitude"), 
                          crs=4326)

# class of sf_ab_listing
class(sf_ab_listing)

# open data 
View(sf_ab_listing)
```

#### 5.3.4 Mapping  

**MAP1**: Quick interactive mapping of Airbnb listings in Denver. 
```{r, eval=FALSE, class.source = "nocopy"}
# set mode to "view" for interactive mapping 
tmap_mode("view")

# MAP1: basic map of Airbnb listing in denver   
tmap_mode("view")
tm_shape(sf_ab_listing) + tm_dots()
```

**[Q9]**: Run the chunk of code below and try to explain the purposes of the following commands: 

* ``` tm_dots() ``` 
* ``` size = "price" ``` 
* ``` col = "room_type" ``` 
* ``` alpha = 0.7 ``` 
```{r, eval=FALSE, class.source = "nocopy"}
# improve MAP1 
tm_shape(sf_ab_listing) + tm_dots(size = "price", col = "room_type", alpha=0.7)
```


### 5.4 Coordinate Reference System (CRS)

Now we have three simple feature objects including ``` co_place ``` collected from US Census, ``` ab_nghbr ``` and ``` sf_ab_listing ``` from [Airbnb data](https://insideairbnb.com/get-the-data/). Let's check their Coordinate Reference System (CRS) and geometry types. 

**[Q10]**: Please report the CRS and geometry types of:

* ``` co_place ``` 
* ``` ab_nghbr ``` 
* ``` sf_ab_listing ``` 
```{r, eval=FALSE, class.source = "nocopy"}
# Check CRS & geometry types 
co_place$geometry
ab_nghbr$geometry
sf_ab_listing$geometry
```

The **World Geodetic System 1984 (WGS 84)** is a global reference system used for determining positions on the Earth's surface. **North American Datum of 1983 (NAD 83)** is the horizontal and geometric control datum for the United States, Canada, Mexico, and Central America. 

Let's use **NAD 83** as the main CRS for this lab and convert WGS 84 to NAD 83. 
```{r, eval=FALSE, class.source = "nocopy"}
# convert "ab_nghbr" and "sf_ab_listing" from WGS84 to NAD83
ab_nghbr <- st_transform(ab_nghbr, st_crs(co_place))
sf_ab_listing <- st_transform(sf_ab_listing, st_crs(co_place))
```

### 5.5 Viz: Mapping Neighborhoods
#### 5.5.1 Neighborhoods in Denver 

We have loaded the spatial data about neighborhoods in Denver in [Step 5.3.1](https://fuzhen-yin.github.io/uccs_geoviz/lab5.html#531_Libraries__Data). . 

Have a look at the neighborhood data ``` ab_nghbr ```. 
```{r, eval=FALSE, class.source = "nocopy"}
View(ab_nghbr)
```


Let's do a quick map of it. 
***Hint**: the following command will produce an error message (no worries, errors are a natural and inevitable part of coding)*. 

**[Q11]**: What error message pops up in the Console? Please try to interpret it using your own words.  
```{r, eval=FALSE, class.source = "nocopy"}
# quick map 
tm_shape(ab_nghbr) + tm_polygons(alpha = 0, lwd = 1)
```

It seems that the data ``` ab_nghbr ``` contains some invalid polygons. Let's make them valid using the function ``` st_make_valid() ```.  
```{r, eval=FALSE, class.source = "nocopy"}
# make valid
ab_nghbr <- st_make_valid(ab_nghbr)
```

#### 5.5.2 Mapping 

MAP2: neighborhoods in Denver
```{r, eval=FALSE, class.source = "nocopy"}
# set plot mode to static 
tmap_mode("plot")

# map 2
map2_den_nghbr <- tm_shape(ab_nghbr) + tm_polygons(alpha = 0, lwd = 1) +
  tm_compass() + tm_scale_bar(width = 0.15) + 
  tm_text(text = "neighbourhood", size = 0.3)

# export 
pdf("plot/map2_denver_neighborhood.pdf")
print(map2_den_nghbr)
dev.off()
```

MAP3: Overlap three maps together (listing, neighborhoods, Denver city boundary)

**[Q11]**: Provide a screenshot of "map3_denver_airbnb_listing_point.pdf". 
```{r, eval=FALSE, class.source = "nocopy"}
# extract Denver city boundary from "co_place" 
den_city <- co_place %>% filter(NAME == "Denver")

# map 3
map_den_listing <- tm_shape(den_city) + tm_polygons(alpha = 0, lwd = 2) +
  tm_shape(ab_nghbr) + tm_polygons(alpha = 0, lwd = 1) + 
  tm_shape(sf_ab_listing) + tm_dots(size = "price", 
                                    col = "room_type", 
                                    alpha=0.5) +
  tm_compass() + tm_scale_bar(width = 0.15) + 
  tm_layout(legend.position = c("left", "top"), frame = F)

# export 
pdf("plot/map3_denver_airbnb_listing_point.pdf", width = 12, height = 8)
print(map_den_listing)
dev.off()
```

### 5.6 Viz: Airbnb Listings in Neighborhoods

While the data ``` sf_ab_listing ``` tells us where the listings are located, and the data ``` ab_nghbr ``` shows the boundaries of neighborhoods. But they are not enough. Let's say we want to know more about the **spatial distribution** of those listings in terms of: 

> how many listings are located in each neighborhoods and also their mean prices?

This section performs analysis and produces visualizations of **Airbnb listings at the [neighborhood level]{.ul}**. 


#### 5.6.1 Spatial Joining 

Examine the ``` sf_ab_listing ``` data. 
```{r, eval=FALSE, class.source = "nocopy"}
# count columns in "sf_ab_listing"
ncol(sf_ab_listing)
```

Spatial joining two data  ``` sf_ab_listing ``` and ``` ab_nghbr ``` using the function ``` st_join() ``` as long as the geometries intersect. Save the joined data as a new variable ``` sf_ab_listing_nghbr ```. 
```{r, eval=FALSE, class.source = "nocopy"}
# spatial join to the left of the object 
sf_ab_listing_nghbr <- st_join(sf_ab_listing, ab_nghbr, 
                               left = T, join = st_intersects)
```

After spatial joining, the new data ``` sf_ab_listing_nghbr ``` has a new column compared to ``` sf_ab_listing ```. Please compare those two data and answer **[Q12]**: What is the new column has been added to ``` sf_ab_listing_nghbr ```? 
```{r, eval=FALSE, class.source = "nocopy"}
# count columns in the new variable "sf_ab_listing_nghbr" 
ncol(sf_ab_listing_nghbr)

# open data 
View(sf_ab_listing_nghbr)
```

#### 5.6.2 Analysis at Neighborhood Level









