---
title: "Chapter 7: Network Visualization (1) Flow & Mobility"
output: 
  html_document:
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 5
    toc_float: 
      collapsed: true
css: website.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 7.1 Lab Goals

In this chapter, you will learn

* **clean** & **organize** network data
* create network based on **nodes** and **edges** 
* **map** network


### 7.2 Good Practice

#### 7.2.1 Organizing Folders & Sub-folders
Under the course folder, please create a folder called "**lab7**". Next, in the **lab7** folder, please create two [sub-folders]{.ul} that one is called "**data**" and another one is "**plot**". 

#### 7.2.2 Data
This chapter explores the [Longitudinal Employer-Household Dynamics](https://lehd.ces.census.gov/data/) data from US Census. Particularly, we will play with the [Job-to-Job Flows (J2J)](https://j2jexplorer.ces.census.gov/). 

> **Job-to-Job Flows (J2J)** is a set of statistics on job mobility in the United States. J2J include statistics on: 
>
  * the job-to-job transition rate, 
  * hires and separations to and from employment,
  * earnings changes due to job change, and
  * characteristics of origin and destination jobs for job-to-job transitions. 
> 
These statistics are available at the national, state, and metropolitan area levels and by worker and firm characteristics.


:::: {.bluebox data-latex=""}
::: {.left data-latex=""}
[**Heads-Up! **]{.ul}
:::

You would need to use [Job-to-Job Flows (J2J)](https://j2jexplorer.ces.census.gov/) for Assignment 3. 

::::


Please follow the steps below to [download]{.ul} data, [unzip]{.ul} it and [move]{.ul} the data to the required folder. 

* **Go** to https://github.com/fuzhen-yin/uccs_geoviz/blob/main/data/lab7_data.zip
* **Download** the file _"lab7_data.zip"_
* **Unzip** folder _"lab7_data.zip"_
* **Move** all files from the "_lab7_data_" folder to the _"data"_ folder under "_lab7_" see [Step 7.2.1](https://fuzhen-yin.github.io/uccs_geoviz/lab7.html#721_Organizing_Folders__Sub-folders) 


If there you have any questions about the above-mentioned steps, please refer to [Chapter 3.2.3](https://fuzhen-yin.github.io/uccs_geoviz/lab3.html#322_Data) for detailed instructions. 


#### 7.2.3 Launching R Studio

Again, we would like to start a **new project** from scratch with a **clean R Script**. Please do the following steps. If you have any questions about these steps, please refer to the relevant chapters for help. 

* Step 1: Make sure all existing R projects are **properly closed**. 
    + If *not*, please close it by going to *File* --> *Close Project* --> *Save* changes (see [Chapter 2.5](https://fuzhen-yin.github.io/uccs_geoviz/lab2.html#25_Close__Exit)). 
* Step 2: Create a **New Project** using **Existing Directory**, navigate to **lab7**, click **open**, then **Create Project**. (see [Chapter 1.3](https://fuzhen-yin.github.io/uccs_geoviz/lab1.html#13_Launching_RStudio)). 
* Step 3: Create a **New Script** by go to **File** --> **New File** --> **R Script**. **Save** the script by giving it a proper name.


#### 7.2.4 Before Start 

:::: {.bluebox data-latex=""}
::: {.left data-latex=""}
[**Heads-Up! **]{.ul}
:::

All scripts are **non-copyable**. 
Lab questions are at the end of this tutorial. 

::::

<br>


### 7.3 Library & Data

#### 7.3.1 Load Libraries
```{r, class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# Library ----
library(dplyr)
library(data.table)
library(sf)
library(tidyverse)
library(forcats)    # Reorder factor levels
library(igraph)     # network object
library(randomcoloR) # random color generator
```

#### 7.3.2 Data
```{r, class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# US Census Cartographic Boundaries: US States (simple features)
sf_us_state <- st_read("data/cb_2023_us_state/cb_2023_us_state_5m.shp")

# Job-to-job flow (j2j) ----
### metadata: request info
j2j_meta_info <- read.csv("data/job_to_job_FROM_Colorado/request_info.csv")
### metadata: state
j2j_meta_state <- read.csv("data/job_to_job_FROM_Colorado/labels_state.csv")
### metadata: North American Industry Classification System (NAICS) code 
j2j_meta_naics <- read.csv("data/job_to_job_FROM_Colorado/labels_naics2_orig.csv")

### flows: separations from COLORADO 
j2j_from_co <- read.csv("data/job_to_job_FROM_Colorado/table.csv")
### flows: hires to COLORADO
j2j_to_co <- read.csv("data/job_to_job_TO_Colorado/table.csv")

```

### 7.4 J2J Viz: J2J Flow by Industry

#### 7.4.1 Metadata
```{r, class.source = "nocopy", eval=FALSE,  message=FALSE, warning=FALSE, cache=FALSE}
### metadata of the request info. 
### scroll down. flags, Year/Quarter: 2023 Q1; 2023 Q2; 2023 Q3
View(j2j_meta_info)
### metadata of state
View(j2j_meta_state)
### metadata of NAICS Sectors 
View(j2j_meta_naics)
```

#### 7.4.2 J2J Job Outflow
```{r, class.source = "nocopy", eval=FALSE,  message=FALSE, warning=FALSE, cache=FALSE}
### j2j_from_co: What states do workers move to? search '.flag'
View(j2j_from_co)
### data frame shape: 20 obs. 101 variables
str(j2j_from_co)
```

##### 7.4.2.1 Tidy Data
```{r, class.source = "nocopy"}
### tidy the data 
j2j_from_co_cln <- j2j_from_co %>% 
  pivot_longer(cols = 2:ncol(j2j_from_co), 
               names_to = "to_state", values_to = "avg_n_job")
```


```{r, class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
#### remove the record with ".flag". "!";    %like% operator.  find names containing the pattern ".flag"
j2j_from_co_cln <- j2j_from_co_cln %>% 
  filter(!to_state %like% ".flag")

#### remove missing values (na - not available) 
j2j_from_co_cln <- j2j_from_co_cln %>% na.omit()

#### update first column name from "X" to "naics_sector"
setnames(j2j_from_co_cln, old="X", new="naics_sector")

#### add a new column indicating the origin state of job flow
j2j_from_co_cln$from_state <- "Colorado"

#### add a new column showing flow director 
j2j_from_co_cln$direction <- "outflow_from_co"

#### correct state names by replacing dot "." with space " "
j2j_from_co_cln$to_state <- gsub("\\.", " " ,j2j_from_co_cln$to_state )
```


```{r, class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
#### re-order columns
j2j_from_co_cln <- j2j_from_co_cln[c("from_state", "to_state", 
                                     "direction", "naics_sector", "avg_n_job")]
```

##### 7.4.2.2 Quick Plot 
plot: job outflow from Colorado by industry 
```{r, class.source = "nocopy", }

##### data preparation
data <- j2j_from_co_cln 

#### box plot, reorder by median
ggplot(data = data, aes(x = reorder(naics_sector, avg_n_job), 
                        y=avg_n_job, fill =naics_sector )) + 
  geom_boxplot() + 
  xlab("NAICS Sector") +
  ylab("Job Counts") + 
  ggtitle("2023 Q1-Q3 Quaterly Average Job Outflow from Colorado") +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 30, vjust = 1, hjust=1)) 
```


#### 7.4.3 J2J Job Inflow
##### 7.4.3.1 Tidy & Cleaning 
Use the pipe operator ``` %>% ``` to chain functions together. 
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
### tidy data; remove records with ".flag" suffix, remove na
j2j_to_co_cln <-  j2j_to_co %>% 
  pivot_longer(cols = 2:ncol(.), 
               names_to = "naics_sector", values_to = "avg_n_job" ) %>% 
  filter(!naics_sector %like% ".flag") %>% 
  na.omit() %>% 
  setnames(old="X", new="from_state") %>% 
  mutate(to_state = "Colorado", direction="inflow_to_co") %>% 
  select(from_state, to_state, direction, naics_sector, avg_n_job)
```

Correct naics names. **Hint**: the code below is **copyable**. 
```{r, message=FALSE, warning=FALSE, cache=FALSE}
j2j_to_co_cln$naics_sector <- j2j_to_co_cln$naics_sector %>% 
  gsub("\\.\\.", ", ", .) %>% 
  gsub("\\.", " ", .) %>% 
  gsub("Other Services, except Public Administration", 
       "Other Services (except Public Administration)", .) %>% 
  gsub("^\\s+|\\s+$", "", .)
```

##### 7.4.3.2 Box Plot of Job Inflow by Industry
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
##### data preparation
data <- j2j_to_co_cln 

#### box plot, reorder by median
ggplot(data = data, 
       aes(x = reorder(naics_sector, avg_n_job), 
           y=avg_n_job, 
           fill =naics_sector )) + 
  geom_boxplot() + 
  xlab("NAICS Sector") +
  ylab("Job Counts") + 
  ggtitle("2023 Q1-Q3 Quaterly Average Job Inflow to Colorado") +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 30, vjust = 1, hjust=1)) 

```

### 7.5 J2J Viz: Network

#### 7.5.1 Data Preparation 
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
### combine inflow & outflow dataframe together
j2j_co_bysector <- rbind(j2j_from_co_cln, j2j_to_co_cln)
### combine all sectors together 
j2j_co_aggsector <- aggregate(avg_n_job ~ from_state + to_state + direction, 
                        data = j2j_co_bysector %>% select(-naics_sector), 
                        FUN = sum)
j2j_co_aggsector$naics_sector <- "all_sectors"


### Final data
j2j_co <- rbind(j2j_co_aggsector, j2j_co_bysector)
```


```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
### list of unique states in j2j_co
lt_state <- unique(c(j2j_co$from_state, j2j_co$to_state))
### list of unique naics sectors in j2j_co
lt_sector <- unique(j2j_co$naics_sector)
```

```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
### visualize job inflow network
data <- j2j_co %>% filter(naics_sector == "all_sectors", 
                          direction == "inflow_to_co")

# Create a network object 
network <- graph_from_data_frame(d=data, directed=F) 

# Network summary 
summary(network)
```

#### 7.5.2 Network Visualization 
Quick plot of network 
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
plot(network)
```

Improve network viz. 
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
plot(network, vertex.size = 12, 
     edge.arrow.size = 0.5,
     edge.width=E(network)$avg_n_job * 0.002, 
     edge.curved = TRUE, main = "Job inflow to Colorado")
```

### 7.6 Viz: Mapping J2J Network

#### 7.6.1 Data Preparation
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
#### project to sf_us_state to CRS NAD83 / Colorado North (ftUS)
sf_us_state_j2j <- sf_us_state %>%
  filter(NAME %in% lt_state) %>% st_transform(., 2231)
```


```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
#### get state centroid  
df_node <- sf_us_state_j2j %>% select(NAME) %>% st_centroid()
#### split point data into lat & long 
df_node <- df_node %>% mutate( long = unlist(map(df_node$geometry, 1)),
                               lat = unlist(map(df_node$geometry, 2)))
st_geometry(df_node) <- NULL
```


```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
#### join coordinates to edge 
j2j_co_xy <- j2j_co %>% 
  left_join(., df_node, by=c("from_state"="NAME")) %>% 
  setnames(old=c("long","lat"), new=c("long_start","lat_start")) %>% 
  left_join(., df_node, by=c("to_state"="NAME")) %>% 
  setnames(old=c("long","lat"), new=c("long_end","lat_end"))

```

#### 7.6.2 Mapping Job Inflow Network 

```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
#### filter data 
data <- j2j_co_xy %>% filter(direction == "inflow_to_co", 
                             naics_sector == "all_sectors")

#### plot 
ggplot(sf_us_state_j2j) + 
  geom_sf() + 
  geom_segment(data = data,
               aes(x = long_start, y = lat_start, 
                   xend = long_end, yend = lat_end, 
                   size = avg_n_job, alpha = 0.5)) + 
  scale_size_continuous(range = c(0.1, 3)) + 
  geom_point(data =df_node , aes(x = long, y = lat), 
             shape = 18, fill = "white",
             color = 'black', stroke = 0.5) + 
  geom_text(data =df_node, 
            aes(x = long, y = lat, label = NAME), 
            size=2) + 
  ggtitle("Job Inflow to Colorado") + 
    ylab("Latitude") + 
    xlab("Longitude")
```

#### 7.6.2 Function: Job Outflow Network 
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE}
### create a function to plot job outflow by sectors  
map_j2j <- function(idx, full_data, state_boundary, state_point, 
                    c_direction, c_sector, edge_color){
  data <- full_data %>% filter(direction == as.character(c_direction),
                               naics_sector == as.character(c_sector))
  title <- paste("Job-to-job", as.character(c_direction), 
                 as.character(c_sector), sep=" ")
  p <- ggplot(state_boundary) + 
    geom_sf() + 
    geom_segment(data = data,
                 aes(x = long_start, y = lat_start, 
                     xend = long_end, yend = lat_end, 
                     size = avg_n_job, alpha = 0.5), 
                 colour = edge_color) + 
    scale_size_continuous(range = c(0.01, 4)) + 
    geom_point(data =state_point , aes(x = long, y = lat), 
               shape = 18, fill = "white",
               color = 'black', stroke = 0.5) + 
    geom_text(data =state_point, 
              aes(x = long, y = lat, label = NAME), 
              size=2) + 
    ggtitle(title) + 
    theme_bw() + 
    ylab("Latitude") + 
    xlab("Longitude")
  
  pdf(sprintf("plot/p%s_%s.pdf", idx, title),  width = 10, height = 5)
  print(p)
  dev.off()
}
```

Mapping job outflow by industry
```{r, class.source = "nocopy", message=FALSE, warning=FALSE, cache=FALSE,  eval=FALSE}
### random color set 
n <- length(lt_sector)
col_random <- randomColor(n)


### map 
for (i in 1:length(lt_sector)) {
  edge_color = col_random[[i]]
  sector_toviz <- lt_sector[[i]]
  
  map_j2j(i, j2j_co_xy, sf_us_state_j2j, df_node, "outflow_from_co", 
          sector_toviz, edge_color)
}

```


### 7.7 Lab Questions 

**[Q1]** Please list **3-5 functions** that you have learnt from this tutorial and explain when and how to use them. **Hint**: a function is an object followed by ``` () ```.  For example, ``` left_join() ```, ``` str() ``` are functions. 

**[Q2]** Please open the _*plot*_ folder, you should be able to see 21 plots. Please pick **2 plots** to discuss your findings. **Hint**: the data frame ``` j2j_co ``` contains the original data of the plots. 

**[Q3]** Have a look at the map **p1_Job-to-job outflow_from_co all_sectors.pdf**, how would you like to improve it? 





### 7.7 Close & Exit

Congratulations!! You have completed the entire tutorial and learnt the intro to network visualization!! Excellent work. 

Please go "File"--> "Close Project" -- a pop window asking "Do you want to save these changes" --> "Yes".

Don't forget to submit the **lab6 report**  and your **script**  to Canvas. 






