---
title: "Chapter 9: Mapping Point PatternS"
output: 
  html_document:
    fig_caption: yes
    theme: cosmo
    toc: yes
    toc_depth: 5
    toc_float: 
      collapsed: true
css: website.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 9.1 Lab Goals

This chapter explores mapping of point patterns:

* Crating **grids** covering study area 
* **Point pattern** visualization 
* **Kernel Density Estimation** Mapping


### 9.2 Good Practice

#### 9.2.1 Organizing Folders & Sub-folders
Under the course folder, please create a folder called "**lab9**". Next, in the **lab9** folder, please create two [sub-folders]{.ul} that one is called "**data**" and another one is "**plot**". 

#### 9.2.2 Data
This chapter uses:

* [Crash data](https://policedata.coloradosprings.gov/Traffic-Crashes/Crash-Data/bjpt-tkzq/about_data) reported to CSPD in 2024.
* 2023 [Tiger Line Road Data](https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2023&layergroup=Roads) from US Census Bureau
* 2023 Cartographic Boundary [County Subdivision](https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html) from US Census Bureau


Download the lab's data: 

* **Go** to https://github.com/fuzhen-yin/uccs_geoviz/blob/main/data/lab9_data.zip
* **Download** the file _"lab9_data.zip"_
* **Unzip** folder _"lab9_data.zip"_
* **Move** all files from the "_lab9_data_" folder to the _"data"_ folder under "_lab9_" see [Step 9.2.1](https://fuzhen-yin.github.io/uccs_geoviz/lab9.html#921_Organizing_Folders__Sub-folders) 


If there you have any questions about the above-mentioned steps, please refer to [Chapter 3.2.3](https://fuzhen-yin.github.io/uccs_geoviz/lab3.html#322_Data) for detailed instructions. 


#### 9.2.3 Launching R Studio

Again, we would like to start a **new project** from scratch with a **clean R Script**. Please do the following steps. If you have any questions about these steps, please refer to the relevant chapters for help. 

* Step 1: Make sure all existing R projects are **properly closed**. 
    + If *not*, please close it by going to *File* --> *Close Project* --> *Save* changes (see [Chapter 2.5](https://fuzhen-yin.github.io/uccs_geoviz/lab2.html#25_Close__Exit)). 
* Step 2: Create a **New Project** using **Existing Directory**, navigate to **lab9**, click **open**, then **Create Project**. (see [Chapter 1.3](https://fuzhen-yin.github.io/uccs_geoviz/lab1.html#13_Launching_RStudio)). 
* Step 3: Create a **New Script** by go to **File** --> **New File** --> **R Script**. **Save** the script by giving it a proper name.


### 9.3 Libraries & Data 

#### 9.3.1 Load Libraries
Please install necessary packages as needed.  

```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
library(dplyr)    # mutate, left_join
library(sf)
library(geojsonio)  ## import geojson
library(tmap)       ## mapping
library(spatstat)   ## kernel density map
```

#### 9.3.2 Data
Simple Features (sf) - spatial data 
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# colorado state: county subdivision (cartographic boundary)
sf_co_cousub <- st_read("data/cb_2023_us_cousub/cb_2023_us_cousub_500k.shp") %>% 
  filter(STATE_NAME == "Colorado")

# crash data
sf_crash <- st_read("data/crash_data/cos_crash_2024.shp")

# road data 
sf_road <- st_read("data/tl_2023_08041_roads/tl_2023_08041_roads.shp")
```

### 9.4 Data Processing 
#### 9.4.1 Spatial Boundaries

Project to CRS NAD83 / Colorado North (ftUS). [CRS code: 2231](https://epsg.io/2231)
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# project to CRS NAD83 / Colorado North (ftUS) 
sf_co_cousub <- st_transform(sf_co_cousub, 2231)
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# el paso county: county subdivision 
sf_ep_cousub <- sf_co_cousub %>% filter(NAMELSADCO == "El Paso County")
# olorado springs 
sf_cos_cousub <- sf_ep_cousub %>% filter(NAME == "Colorado Springs")
```

#### 9.4.2 Crash Data

Examine the crash data. 
```{r, eval=FALSE, class.source = "nocopy"}
# top 5 rows 
head(sf_crash, 5)
# data type 
str(sf_crash)
```

Filter data 
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# extract records with non-empty geometry 
sf_crash_24 <- sf_crash[!st_is_empty(sf_crash),]

# project to CRS NAD83 / Colorado North (ftUS)
sf_crash_24 <- st_transform(sf_crash_24, 2231)
```

#### 9.4.3 Exploring Analysis

Visualization: quick interactive map.
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# set tmap mode for interactive mapping 
tmap_mode("view")

# plot crash data and cos county subdivision
tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, lwd = 5, border.col = "blue") + 
  tm_shape(sf_crash_24) + tm_dots(alpha = 0.5)
```

Here are some issues with the crash data, (1) many dots overlap with each other; (2) there are crash dots outside of cos city boundary. Let's extract only crash within OS city boundary
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# extract only crash within OS city boundary 
sf_crash_24_cos <- sf_crash_24[lengths(st_intersects(sf_crash_24, sf_cos_cousub))>0, ]
```


### 9.5 Viz: Grid Point Pattern

#### 9.5.1 Grids 

```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# create a grid cover the entire city by defining the cell size (each grid will be 2000 ft * 2000 ft * )
sf_grid_cos <- st_make_grid(sf_cos_cousub, cellsize =  2000, square = TRUE ) %>% st_as_sf() 

# visualize the grid 
tm_shape(sf_grid_cos) + tm_polygons() + 
  tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, border.col = "red")
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# remove grid outside of COS boundary 
sf_grid_cos <- sf_grid_cos[lengths(st_intersects(sf_grid_cos, sf_cos_cousub)) > 0, ]

# view 
tm_shape(sf_grid_cos) + tm_polygons() + 
  tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, border.col = "red")
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# count crashes in each grid 
sf_grid_cos$n_crash <- lengths(st_intersects(sf_grid_cos, sf_crash_24_cos))

# view 
tm_shape(sf_grid_cos) + tm_polygons(col = "n_crash", border.alpha = 0.2) + 
  tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, border.col = "red") 
```

#### 9.5.2 Roads
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# projection 
sf_road <- st_transform(sf_road, 2231)

# extract main roads
# MTFCC code: https://www2.census.gov/geo/pdfs/reference/mtfccs2022.pdf
sf_r_main <- sf_road %>% filter(MTFCC %in% c("S1100", "S1200")) 
sf_r_main <- sf_r_main[lengths(st_intersects(sf_r_main, sf_grid_cos)) > 0, ]

# prepare labels 
sf_r_main_lbl <- sf_r_main[!duplicated(sf_r_main$FULLNAME),]
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
## Add main roads to the grid point map
tm_shape(sf_grid_cos) + tm_polygons(col = "n_crash", border.alpha = 0.2) + 
  tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, border.col = "red") + 
  tm_shape(sf_r_main) + tm_lines() + 
  tm_shape(sf_r_main_lbl) + tm_text(text = "FULLNAME", size = 0.7, remove.overlap = T)
```

#### 9.5.3 Crash Points by Vehicle Types 

```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# vehicle types 
lt_v_type <- unique(sf_crash_24_cos$vehicletyp) 

# remove NA
lt_v_type <- lt_v_type[!is.na(lt_v_type)]

# lbl 
lt_v_type_lbl <- c("auto", "motor", "truck", "bus", "bike", "trailor")

print(lt_v_type)
print(lt_v_type_lbl)
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# county the number of crash by vehicle types in each grid  
for (i in 1:length(lt_v_type)){
  print(i)
  print(lt_v_type[[i]])
  
  data <- sf_crash_24_cos %>% filter(vehicletyp == lt_v_type[[i]] )
  sf_grid_cos[as.character(lt_v_type_lbl[[i]])] <- lengths(st_intersects(sf_grid_cos, 
                                                                         data))
}

# check the output data 
head(sf_grid_cos, 10)
```

Map crash involving motors. 

```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
tmap_mode("plot")

tm_shape(sf_grid_cos) + tm_polygons(col = "motor", border.alpha = 0.2, palette = "Blues") + 
  tm_shape(sf_cos_cousub) + tm_polygons(alpha = 0, border.col = "red") + 
  tm_shape(sf_r_main) + tm_lines() + 
  tm_shape(sf_r_main_lbl) + tm_text(text = "FULLNAME", size = 0.5, remove.overlap = T) + 
  tm_compass(north = 0, size = 1, position = c("right", "top")) + 
  tm_scale_bar(position = c("right", "bottom"), width = 0.15)  + 
  tm_layout(legend.outside = T)
```


:::: {.bluebox data-latex=""}
::: {.left data-latex=""}
[**Q1 **]{.ul}
:::
Please adapt the code chunk above to produce the following maps, and paste the screenshots to the report. 

 * crashes involving automobiles using the color palette "PuBuGn"
 * crashes involving trailers using the color palette "YlOrRd"
 * crashes involving buses using the color palette "Greens"

::::



### 9.6 Viz: Kernel Density Map

**Hint** You would need to re-write this section (9.6) to answer **Q2**. 

**Q2**: Please produce a kernel density map for crash involving trucks, and paste the screenshot to the report.

#### 9.6.1 Data Preparation
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# data
# spatial polygon of city boundary: sf_cos_cousub 
# spatial point of crash data: sf_crash_24_cos

## check different vehicle types 
print(lt_v_type)
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
## filter crash of automobile 
data <- sf_crash_24_cos %>% filter(vehicletyp == "Automobile")
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# check the class of "data"
class(data)

# convert from simple features to spatial point 
sp_data <- as(data, "Spatial")

# check class "sp_data"
class(sp_data)
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# coordinates top 10 rows 
head(sp_data@coords, 10)
```

#### 9.6.2 KDE Mapping
```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# create observation window (owin)
wndw_cos <- as.owin(sf_cos_cousub)
plot(wndw_cos)

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
#create a ppp (point pattern) object
ppp_crash_auto_24_cos <- ppp(x=sp_data@coords[,1],
                        y=sp_data@coords[,2],
                        window=wndw_cos)
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# kernel density map
ppp_crash_auto_24_cos %>%
  density(., sigma=500) %>%
  plot()
```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}
# play with sigma 
ppp_crash_auto_24_cos %>%
  density(., sigma=2000) %>%
  plot()
```

### 9.7 Close & Exit

Congratulations!! You have completed the entire tutorial and learnt the intro to network visualization!! 

**Please submit your report on time, otherwise, a late penalty (5 pts / day) will be applied**. 

Please go "File"--> "Close Project" -- a pop window asking "Do you want to save these changes" --> "Yes".

Don't forget to submit the **lab9 report**  and your **script**  to Canvas. 

```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```


```{r class.source = "nocopy",  message=FALSE, warning=FALSE, cache=FALSE}

```










